---
title: "Student Growth and District Demographics Analysis"
author: "MCAS Data Analysis"
date: "`r Sys.Date()`"
format: 
  typst:
    toc: true
    number_sections: true
    highlight_style: github
    mainfont: "Arial"
    fontsize: 11pt
    geometry: "margin=1in"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                     fig.width = 10, fig.height = 7, fig.align = "center",
                     out.width = "100%", dpi = 300)
```

## Overview

This analysis examines the relationship between district racial composition (proportion of white students) and student academic growth as measured by Student Growth Percentiles (SGP) on Massachusetts MCAS tests. We use weighted regression where each district is weighted by its total number of students.

## Libraries and Data Loading

```{r libraries}
library(RSQLite)
library(dplyr)
library(ggplot2)
library(estimatr)  # for robust standard errors in weighted regression
library(broom)
library(knitr)
library(tidyr)
library(scales)

# Subject labels for display
subject_labels <- c("ELA" = "English", "MATH" = "Math", "SCI" = "Science")
```

```{r load-data}
# Connect to SQLite database
con <- dbConnect(SQLite(), "../../data/processed/mcas.db")

# Read MCAS data
query <- "
  SELECT 
    DIST_NAME as district,
    year,
    SUBJECT_CODE as subject,
    AVG(avg_sgp) as avg_sgp,
    SUM(n_white) as total_n_white,
    SUM(n) as total_n,
    CASE 
      WHEN SUM(n) > 0 THEN CAST(SUM(n_white) AS FLOAT) / SUM(n)
      ELSE NULL 
    END as prop_white
  FROM mcas 
  WHERE SUBSTR(ORG_CODE, -4) = '0000' 
    AND avg_sgp IS NOT NULL
    AND n > 0
    AND n_white IS NOT NULL
    AND DIST_NAME != 'State'
  GROUP BY DIST_NAME, year, SUBJECT_CODE
  HAVING SUM(n) > 100
  ORDER BY district, year, subject
"

mcas_data <- dbGetQuery(con, query)
dbDisconnect(con)

# Filter out null values and ensure reasonable bounds
mcas_data <- mcas_data %>%
  filter(
    !is.na(avg_sgp),
    !is.na(prop_white),
    prop_white >= 0,
    prop_white <= 1
  )

cat("Data loaded:", nrow(mcas_data), "observations\n")
cat("Years:", min(mcas_data$year), "-", max(mcas_data$year), "\n")
cat("Districts:", length(unique(mcas_data$district)), "\n")
cat("Subjects:", paste(unique(mcas_data$subject), collapse = ", "), "\n")
```

## Weighted Regression Analysis

```{r regression-analysis}
# Function to run weighted regression with robust standard errors
run_weighted_regression <- function(data) {
  if (nrow(data) < 3) return(NULL)
  
  # Use estimatr::lm_robust for weighted regression with robust standard errors
  model <- lm_robust(avg_sgp ~ prop_white, 
                     data = data, 
                     weights = total_n,
                     se_type = "HC2")  # Heteroskedasticity-consistent standard errors
  
  # Extract coefficients
  coefs <- tidy(model)
  slope_row <- coefs[coefs$term == "prop_white", ]
  
  return(data.frame(
    slope = slope_row$estimate,
    std_error = slope_row$std.error,
    t_stat = slope_row$statistic,
    p_value = slope_row$p.value,
    conf_low = slope_row$conf.low,
    conf_high = slope_row$conf.high,
    n_districts = nrow(data),
    total_students = sum(data$total_n)
  ))
}

# Run regressions for each year-subject combination
regression_results <- mcas_data %>%
  group_by(year, subject) %>%
  do(run_weighted_regression(.)) %>%
  filter(!is.na(slope)) %>%
  mutate(
    subject_label = recode(subject, !!!subject_labels),
    significant = p_value < 0.05
  )

cat("Regression results for", nrow(regression_results), "year-subject combinations\n")
```

## Regression Results Table

```{r results-table}
# Create formatted table
results_table <- regression_results %>%
  ungroup() %>%  # Remove grouping to avoid column mismatch
  select(year, subject_label, slope, std_error, t_stat, p_value, n_districts, total_students) %>%
  mutate(
    slope = round(slope, 3),
    std_error = round(std_error, 3),
    t_stat = round(t_stat, 2),
    p_value = case_when(
      p_value < 0.001 ~ "< 0.001",
      p_value < 0.01 ~ "< 0.01", 
      p_value < 0.05 ~ "< 0.05",
      TRUE ~ as.character(round(p_value, 3))
    ),
    total_students = scales::comma(total_students)
  )

kable(results_table, 
      col.names = c("Year", "Subject", "Slope", "Std Error", "t-stat", "p-value", "# Districts", "Total Students"),
      caption = "Weighted Regression Results: Student Growth vs. Proportion White Students",
      align = c("c", "l", "r", "r", "r", "r", "r", "r"))
```

## Key Findings

```{r summary-stats}
# Summary statistics
avg_slope <- mean(regression_results$slope)
significant_results <- sum(regression_results$significant)
negative_slopes <- sum(regression_results$slope < 0)

cat("Average slope across all year-subject combinations:", round(avg_slope, 3), "\n")
cat("Statistically significant results (p < 0.05):", significant_results, "out of", nrow(regression_results), "\n")
cat("Negative slope estimates:", negative_slopes, "out of", nrow(regression_results), "\n")
```

## Visualization

```{r main-plot, fig.width=12, fig.height=9}
# Create the main scatter plot with facets
p1 <- ggplot(mcas_data, aes(x = prop_white, y = avg_sgp)) +
  geom_point(aes(size = total_n), alpha = 0.6, color = "#2563eb") +
  geom_smooth(aes(weight = total_n), method = "lm", se = TRUE, color = "#dc2626", size = 1) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "#666666", alpha = 0.7) +
  facet_grid(year ~ subject, labeller = labeller(subject = subject_labels)) +
  scale_x_continuous(labels = percent_format(), name = "Proportion of White Students") +
  scale_y_continuous(name = "Average Student Growth Percentile") +
  scale_size_continuous(name = "Total Students", 
                       range = c(1, 8),
                       labels = comma_format()) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    axis.title = element_text(size = 12),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Student Growth vs. Proportion of White Students by Year and Subject",
    subtitle = "Each point represents a school district, sized by number of students. Red lines show weighted regression trends.",
    caption = "Data: Massachusetts MCAS. Districts weighted by student enrollment. Dashed line at 50th percentile (typical growth)."
  )

print(p1)
```

## Slope Trends Over Time

```{r slope-trends, fig.width=10, fig.height=5}
# Plot slope estimates over time by subject
p2 <- ggplot(regression_results, aes(x = year, y = slope, color = subject_label)) +
  geom_line(size = 1.2) +
  geom_point(aes(size = total_students), alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", alpha = 0.5) +
  scale_color_brewer(type = "qual", palette = "Set1", name = "Subject") +
  scale_size_continuous(name = "Total Students", 
                       labels = comma_format(),
                       range = c(2, 6)) +
  scale_x_continuous(breaks = pretty_breaks()) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor.x = element_blank()
  ) +
  labs(
    title = "Regression Slope Estimates Over Time",
    subtitle = "Change in student growth percentile per unit increase in proportion white students",
    x = "Year",
    y = "Slope Estimate",
    caption = "Points sized by total student enrollment across districts"
  )

print(p2)
```

## Statistical Significance Over Time

```{r significance-plot, fig.width=10, fig.height=5}
# Plot significance over time
p3 <- regression_results %>%
  mutate(significance_level = case_when(
    p_value < 0.001 ~ "p < 0.001",
    p_value < 0.01 ~ "p < 0.01",
    p_value < 0.05 ~ "p < 0.05",
    TRUE ~ "p ≥ 0.05"
  )) %>%
  ggplot(aes(x = year, y = slope, color = significance_level, shape = subject_label)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", alpha = 0.5) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(
    values = c("p < 0.001" = "#d73027", "p < 0.01" = "#fc8d59", 
               "p < 0.05" = "#fee08b", "p ≥ 0.05" = "#999999"),
    name = "Statistical Significance"
  ) +
  scale_shape_manual(values = c(16, 17, 15), name = "Subject") +
  scale_x_continuous(breaks = pretty_breaks()) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(
    title = "Statistical Significance of Slope Estimates",
    subtitle = "Relationship between proportion white students and student growth",
    x = "Year",
    y = "Slope Estimate",
    caption = "Warmer colors indicate stronger statistical significance"
  )

print(p3)
```

## Interpretation

The analysis reveals several key patterns:

1. **Overall Relationship**: On average, there is a `r ifelse(avg_slope > 0, "positive", "negative")` relationship between the proportion of white students in a district and student growth percentiles.

2. **Statistical Significance**: `r significant_results` out of `r nrow(regression_results)` year-subject combinations show statistically significant relationships (p < 0.05).

3. **Variation by Subject and Year**: The strength and direction of the relationship varies across subjects and over time, suggesting complex underlying dynamics.

4. **Weighting Impact**: By weighting districts by student enrollment, larger districts (which affect more students) have greater influence on the estimated relationships.

**Note**: These are observational correlations and should not be interpreted as causal relationships. Many factors beyond racial composition influence student academic growth.
